
---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---

# Reproducible Research: Peer Assessment 1

## Introduction
This is the first peer assessment for the Coursera course Reproducible Research.

#### Required packages
```{r, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(tidyverse)
library(lubridate)
library(rmarkdown)
library(knitr)
```



## Loading and preprocessing the data
#### Load the Data
```{r}
activity <- read.csv(file="activity.csv", header=TRUE)
```
#### Understand the Data
```{r}
head(activity, 2)
str(activity)
```


## What is mean total number of steps taken per day?
#### Transform the data for analysis
```{r}
tot_steps <- activity %>% group_by(date) %>% filter(!is.na(steps)) %>%
      summarize(tot_steps = sum(steps, na.rm = TRUE))
```

#### What is the mean total number of steps taken per day
#### and the median total number of steps taken per day?
```{r}
summary(tot_steps$tot_steps) ##  or 
avg_steps <- mean(tot_steps$tot_steps, na.rm = TRUE)
median_steps <- median(tot_steps$tot_steps, na.rm = TRUE)
```
#### Make a histogram of the total number of steps taken per day
```{r}
ggplot(tot_steps, aes(tot_steps)) + geom_histogram(binwidth = 2500, fill = "darkgreen", col = "lightgreen") +
      labs(title = "Daily Steps Histogram", x = "Total Steps per Day", y = "Number of Days")
```

## What is the average daily activity pattern?
#### Transform the data for analysis
```{r}
step_int <- activity %>% group_by(interval) %>% filter(!is.na(steps)) %>%
      summarize(int_steps = mean(steps, na.rm = TRUE))
head(step_int, 2)
str(step_int)

summary(step_int)
```

#### Make a time series plot of the data
```{r}
ggplot(step_int, aes(interval, int_steps)) + geom_line() +
      labs(title = "Daily Average Activity", x = "Interval", y = "Average Steps")
```


## Interval with most steps
#### Calculate which interval has the highest average steps per day
```{r}
step_int[which.max(step_int$int_steps), ] # 835 which means 8:35-8:40
```

## Imputing missing values
#### Summary of missing values
```{r}
colSums(is.na(activity)) # 2304 missing steps values, 0 missing date values, 0 missing interval values
```

## Add mean value for that interval into original dataset replacing NA values
#### Transform the data for analysis
```{r}
fj_int <- activity
fj_int_steps <- full_join(fj_int, step_int, by = 'interval')
nas <- is.na(fj_int$steps)
fj_int_steps[nas,]$steps <- fj_int[nas, ]$int_steps
filled_int_steps <- select(fj_int_steps, steps = steps, date, interval)
filled_int_steps_date <- filled_int_steps %>% group_by(date)
head(filled_int_steps_date, 5)
colSums(is.na(filled_int_steps_date))

filled_tot_steps <- summarize(filled_int_steps_date, total_steps = sum(steps))
```

#### Plot the daily steps data
```{r}
ggplot(filled_tot_steps, aes(total_steps)) + geom_histogram(binwidth = 2500, fill = "darkblue", col = "lightblue") +
      labs(title = "Daily Steps Histogram", x = "Total Steps per Day", y = "Number of Days")

summary(filled_tot_steps)
```

## Are there differences in activity patterns between weekdays and weekends?
#### Transform the data for analysis
```{r warning = FALSE}
filled_int_steps_date <- mutate(filled_int_steps_date, weekend = factor(weekdays(ymd(date)) %in% c('Saturday', 'Sunday')))
filled_tot_steps_wk <- group_by(filled_int_steps_date, weekend, interval)
filled_tot_steps_wk_int <- summarize(filled_tot_steps_wk, steps = mean(steps))
```

#### Plots for weekend vs weeekday steps
```{r}
gplot1 <- filled_tot_steps_wk_int %>% filter(weekend == TRUE)
ggplot(gplot1, aes(interval, steps)) + geom_line() + labs(title = "Weekends", x = "Average Interval Steps per Day",
                                                          y = "# of Steps")
gplot2 <- filled_tot_steps_wk_int %>% filter(weekend == FALSE)
ggplot(gplot2, aes(interval, steps)) + geom_line() + labs(title = "Weekdays", x = "Average Interval Steps per Day",
                                                          y = "# of Steps")
```









